Categorical Foundations of E-Graphs With Binding

### 2025-07-10 POPL2026/paper/sections/categorical_semantics.tex
mass format

        % \[\begin{tikzcd}
        % 	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
        % 	\end{tikzcd}
        % \]
        \vspace{-1mm}
        \[
        F \dashv U : \catname{SLat} \to \catname{Set}.
        \vspace{-1mm}
        \]
        Moreover, every symmetric monoidal category $\mathbb{C}$ in $\catname{Cat}$ (respectively, $\catname{SLat}\text{--}\catname{Cat}$) is taken to a symmetric monoidal category $\mathbb{C}'$ in $\catname{SLat}\text{--}\catname{Cat}$ (respectively, $\catname{Cat}$) by a 2-functor $\mathcal{F}$ (respectively, $\mathcal{U}$).
    \end{proposition}
In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor, it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{SLat}$-adjunction.
Thus, applying the 2-functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{SLat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
\begin{aligned}
\Lambda_{A,B,C}(f + g) &= \eta;(B \multimap (f + g))\\
                       &= \eta;(B \multimap (f) + B \multimap (g))\\
                       &= \eta;(B \multimap (f)) + \eta;(B \multimap (g))\\
                       &= \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
\end{aligned}
\end{equation}
\label{law:distributivity}

@@

Slotted e-graphs~\cite{slotted-egraphs} extended the traditional signature employed by e-graphs by an explicit construct for binding ($\text{bind}\; \$x\; tc$) and variables (slots).
\begin{align*}
&\text{functional symbols}\; f, g, \ldots\\
&\text{slots}\;  \$x, \$y, \ldots\\
&\text{term children}\; tc  \Coloneqq t \;|\; tc_{1}, \ldots, tc_{n} \;|\; \text{bind}\; \$x\; tc \;|\; \$x\\ 
&\text{terms}\;  \Coloneqq f \;|\; f(tc)
\end{align*}

>>>>

	% \[\begin{tikzcd}
	% 	\catname{SLatt} \arrow[rr, "U", bend left] & \top & \catname{Set} \arrow[ll, "F", bend left]
	% 	\end{tikzcd}
	% \]
	\vspace{-1mm}
	\[
		F \dashv U : \catname{SLat} \to \catname{Set}.
		\vspace{-1mm}
	\]
	Moreover, every symmetric monoidal category $\mathbb{C}$ in $\catname{Cat}$ (respectively, $\catname{SLat}\text{--}\catname{Cat}$) is taken to a symmetric monoidal category $\mathbb{C}'$ in $\catname{SLat}\text{--}\catname{Cat}$ (respectively, $\catname{Cat}$) by a 2-functor $\mathcal{F}$ (respectively, $\mathcal{U}$).
\end{proposition}
In particular, 2-functor $\mathcal{F}$ turns every category $\mathbb{C'} \in \catname{Cat}$ into a free semilattice enriched $\mathbb{C} \in \catname{SLat}\text{--}\catname{Cat}$ by making every hom-set of $\mathbb{C'}$ into a free semilattice on this set.
Being a 2-functor, it also lifts an adjunction in $\catname{Cat}$ $F \dashv G$ into an $\catname{SLat}$-adjunction.
Thus, applying the 2-functor to $\catname{CS}(\Sigma)$ we get a closed symmetric monoidal $\catname{SLat}$-category $\catname{CS}(\Sigma)^{+}$ by lifting the adjunction $(- \otimes B, B \multimap{} -, \eta, \varepsilon)$.
We can similarly derive $\catname{SLat}$-enriched versions of  $\Lambda_{A,B,C}$ and $\textbf{ev}_{A,B}$ morphisms using the lifted adjunction as $\textbf{ev}_{A,B} = \varepsilon_{A}$ and $\Lambda_{A,B,C}(f) = \eta; (B \multimap{} f)$.
The following distributivity law can then be derived, forming the basis of encoding equivalences inside abstractions.
\begin{equation}
	\begin{aligned}
		\Lambda_{A,B,C}(f + g) & = \eta;(B \multimap (f + g))                      \\
		                       & = \eta;(B \multimap (f) + B \multimap (g))        \\
		                       & = \eta;(B \multimap (f)) + \eta;(B \multimap (g)) \\
		                       & = \Lambda_{A,B,C}(f) + \Lambda_{A,B,C}(g)
	\end{aligned}
\end{equation}
\label{law:distributivity}

@@

Slotted e-graphs~\cite{slotted-egraphs} extended the traditional signature employed by e-graphs by an explicit construct for binding ($\text{bind}\; \$x\; tc$) and variables (slots).
	\begin{align*}
		 & \text{functional symbols}\; f, g, \ldots                                                                   \\
		 & \text{slots}\;  \$x, \$y, \ldots                                                                           \\
		 & \text{term children}\; tc  \Coloneqq t \;|\; tc_{1}, \ldots, tc_{n} \;|\; \text{bind}\; \$x\; tc \;|\; \$x \\
		 & \text{terms}\;  \Coloneqq f \;|\; f(tc)
	\end{align*}

### 2025-07-10 POPL2026/paper/sections/combinatorial_semantics.tex
mass format

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete ordered hypergraphs that denote interfaces.
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the category with discrete \textit{ordered} hypergraphs as objects and isomorphism classes of hypergraphs with interfaces as morphisms.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category with the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:
\begin{align*}
n \xrightarrow{f} &\mathcal{F} \xleftarrow{g} m\\
&\times \qquad \quad \Coloneqq n + n' \xrightarrow{f + f'} \mathcal{F} + \mathcal{G} \xleftarrow{g + g'} m + m'\\
n' \xrightarrow{f'} &\mathcal{G} \xleftarrow{g'} m'
\end{align*}
Unit is given by the initial object of $\catname{Hyp}(\Sigma)$.
Symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.

@@

Consider an example in the middle part of Fig.~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{e-color}{E} = \{e_{1}\}$, $\textcolor{closed-color}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\begin{array}{cccc}
  v_{4} \textcolor{e-color}{<^{\mu}} e_{1} & v_{7} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{e-color}{<^{\mu}} e_{1} & v_{8} \textcolor{closed-color}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{e-color}{<^{\mu}} e_{1} & e_{5} \textcolor{closed-color}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{e-color}{<^{\mu}} e_{1} & w_{9} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]

>>>>

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
	\label{def:cspd}
	A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete ordered hypergraphs that denote interfaces.
	Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the category with discrete \textit{ordered} hypergraphs as objects and isomorphism classes of hypergraphs with interfaces as morphisms.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category with the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:
\begin{align*}
	n \xrightarrow{f}   & \mathcal{F} \xleftarrow{g} m                                                                                   \\
	                    & \times \qquad \quad \Coloneqq n + n' \xrightarrow{f + f'} \mathcal{F} + \mathcal{G} \xleftarrow{g + g'} m + m' \\
	n' \xrightarrow{f'} & \mathcal{G} \xleftarrow{g'} m'
\end{align*}
Unit is given by the initial object of $\catname{Hyp}(\Sigma)$.
Symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.

@@

Consider an example in the middle part of Fig.~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{e-color}{E} = \{e_{1}\}$, $\textcolor{closed-color}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
	\begin{array}{cccc}
		v_{4} \textcolor{e-color}{<^{\mu}} e_{1}  & v_{7} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{4} \consistency e_{2}   & v_{4} \not \consistency v_{10} \\
		e_{3} \textcolor{e-color}{<^{\mu}} e_{1}  & v_{8} \textcolor{closed-color}{<^{\mu}} e_{2} & e_{2} \consistency u_{1}   & e_{2} \not \consistency e_{6}  \\
		e_{2} \textcolor{e-color}{<^{\mu}} e_{1}  & e_{5} \textcolor{closed-color}{<^{\mu}} e_{2} & e_{2} \consistency e_{4}   & e_{4} \not \consistency e_{7}  \\
		\ldots                                    & \ldots                                        & \ldots                     & \ldots                         \\
		v_{12} \textcolor{e-color}{<^{\mu}} e_{1} & w_{9} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
	\end{array}
\]
